---
title: "Bioinformatics exam project: Disease subtype discovery using multi-omics data integration"
author: "Alessia Cecere, Alessandro Di Gioacchino"
date: "19 June 2023"
csl: ieee.csl
bibliography: references.bib
output:
  html_notebook:
    toc: yes
    number_sections: yes
    toc_float: yes
    theme: cerulean
    fig_caption: yes
---

Install required packages:

```{r}
if (!require("BiocManager", quietly = TRUE)) {
  install.packages("BiocManager")
}

BiocManager::install("curatedTCGAData")
BiocManager::install("TCGAutils")
BiocManager::install("TCGAbiolinks")

install.packages("SNFtool")
install.packages("caret")
install.packages("cluster")
install.packages("mclustcomp")
```

Download required packages:

```{r}
library("curatedTCGAData")
library("TCGAbiolinks")
library("TCGAutils")
library("SNFtool")
library("caret")
library("cluster")
library("mclustcomp")
```
model.extract()

1. Download the Prostate adenocarcinoma dataset considering three different omics data sources (mRNA,
miRNA and protein expression data). The TCGA code for the dataset is “PRAD”.

```{r}
assays <- c("miRNASeqGene", "RNASeq2Gene", "RPPAArray")

mo <- curatedTCGAData(
  diseaseCode = "PRAD",
  assays = assays,
  version = "2.0.1",
  dry.run = FALSE
)
# removing unused assay
mo <- mo[, , paste0("PRAD", "_", assays, "-20160128")]
mo
```

2. Pre-process the dataset following the same steps we used during lessons. During the filtering by variance,
select the first 100 features having highest variance from each data source.

```{r}
#  retain only primary tumors
primary <- TCGAutils::TCGAsampleSelect(colnames(mo), c("01"))
mo <- mo[, primary, ]

# checking for technical replicates (not present)
check_rep <- anyReplicated(mo)
print(check_rep)

# removing FFPE (formalin-fixed, paraffin-embedded) samples
no_ffpe <- which(as.data.frame(colData(mo))$patient.samples.sample.is_ffpe == "no")
mo <- mo[, no_ffpe, ]

# retrieve samples having all the considered omics
complete <- intersectColumns(mo)
# extract assays in list
complete <- assays(complete)
# obtain matrices samples x features:
complete <- lapply(complete, FUN=t)

# removing columns with at least one missing value
complete[[1]] <- complete[[1]][, colSums(is.na(complete[[1]])) == 0]
complete[[2]] <- complete[[2]][, colSums(is.na(complete[[2]])) == 0]
complete[[3]] <- complete[[3]][, colSums(is.na(complete[[3]])) == 0]

# removing features with near zero variance and retaining top 100 features having higher variance
nf <- 100

for(i in 1:length(complete)){
    idx <- caret::nearZeroVar(complete[[i]])
    message(paste("Removed ", length(idx), "features from", names(complete)[i]))
    if(length(idx) != 0){
        complete[[i]] <- complete[[i]][, -idx];
    }
    if(ncol(complete[[i]]) <= nf) next
    vars <- apply(complete[[i]], 2, var);
    idx <- sort(vars, index.return=TRUE, decreasing = TRUE)$ix;
    complete[[i]] <- complete[[i]][, idx[1:nf]]
}

# standardizationg features using z-score
zscore <- function(data){
    
    zscore_vec <- function(x) { return ((x - mean(x)) / sd(x))}
    data <- apply(data, 2, zscore_vec)
    
    
    return(data)
}

complete <- lapply(complete, zscore);

# renaming samples
for(v in 1:length(complete)){
    rownames(complete[[v]]) <- substr(rownames(complete[[v]]), 1, 12);
}
```

3. Download the disease subtypes (column “Subtype_Integrative” is the one containing the iCluster
molecular subtypes). Note that not all subtypes are available for the set of samples having all the
considered omics data sources, thus you need to retain from the multi-omics dataset only samples
having an associated subtype.


```{r}
# downloading disease subtypes from TCGAbiolinks:
subtypes <- as.data.frame(TCGAbiolinks::PanCancerAtlas_subtypes());
subtypes <- subtypes[subtypes$cancer.type == "PRAD", ];

# retaining only primary solid tumors
subtypes <- subtypes[TCGAutils::TCGAsampleSelect(subtypes$pan.samplesID, "01"), ];

# selecting subtypes samples that are also present in complete
sub_select <- substr(subtypes$pan.samplesID,1,12) %in% rownames(complete[[1]]);
subtypes <- subtypes[sub_select, ]
rownames(subtypes) <- substr(subtypes$pan.samplesID, 1, 12);

# selecting complete samples that are also present in subtypes
sub_select <- rownames((complete[[1]])) %in% rownames(subtypes)
complete[[1]] <- complete[[1]][sub_select, ]
```

4. Check that patients in multi-omics dataset and subtypes are in the same order.

```{r}
identical(rownames(subtypes), rownames(complete[[1]])) # the function returns FALSE, so they are not in the same order

# reordering
subtypes <- subtypes[rownames(complete[[1]]),]

identical(rownames(subtypes), rownames(complete[[1]])) # now the order is the same
```